Appendix A - Alternative DSDM formulations

### 2019-10-22 manuscript/appendix_a.Rmd
add code for evolutionary model; reorder models; spell check

Each section below details a modelling option for a specific component of the DSDM - the observation model, aundance model, or population dynamic model. These components are largely interoperable; a user can pick one modelling option from each category that best fits the ecology of the species of interest and the eoclogical question at hand. The one exception to this is that abundance models using as input the intrinsic growth rate of the species (the Beverton-Holt model in the manuscript example, and the Ricker model introduced here) cannot be used in combination with the population dynamic models that have either dynamic transition matrices or dispersal, since these models do not yield a separate intrinsic growth rate for each site. With these models, the relative abundance is instead calculated directly from the dynamics, so these abundance models are not needed.

@@

For ease of reference, I repeat here the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript. It presents the model in its three parts, each of which can be modified independently. Specification of priors for parameters in the model are not discussed here. See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta.

## Observation model

@@

Given vectors `Lambda` and `y` each with length $N$, the number of sites.

@@

- `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence

@@

Given a 3D array `A` with dimensions $N$x$M$x$M$, where $N$ is the number of sites, and $M$ the number of life stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar. When used with its default settings the function `greta.dynamics::iterate_matrix()` returns, among other quantities, a vector of intrinsic growth rates of the matrices in `A` computed by repeated iteration of the matrix

@@

The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous timestep (columns) to each stage in the present time step (rows). Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult. For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles

@@

Given an $N$x$K$ matrix `X`, length $K$ vectors `beta_S` and `beta_F`, and a scalar `gamma`, where $N$ is the number of sites and $K$ the number of columns (covariates plus an intercept) of the design matrix.

@@

Below I provide alternative observation models for count, and presence-only data. occupancy detection observation models can be implemnted using other greta extension packages, but are byond the scope of this manuscript.

@@

The relative abundance variable $Lambda_i$ is the parameter of the poisson distribution, so a poisson count model is simply given by:

@@

Where count data are overdispersed with respect to the expected counts $\Lambda_i$, this overdispersion can be modelled with a negative binomial distribution as:

@@

- `r`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution
- `Lambda`: a (positive) vector of the relative abundances for each observation
- `y`: a vector of (non-negative) integers encoding count observations

```{r negbin_obs}
p <- r / (r + Lambda)
distribution(y) <- negative_binomial(r, p)
```
Given a positive scalar `r`, and where y is a vector of counts. `r` and `p` correspond to the `size` and `prob` arguments of `stats::dnbinom` in R.

@@

Presence-only data - where records consist of locations where the species has been reported as being present - can be considered as a thinned, inhomogenous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015]. The likelihood for this model comprises an intractable integral over the study area, which can be approximated in a number of ways that reduce the model to having a Poisson observation model. Here I apply the piecewise constant approximation, defininthe observations $y_i$ to be counts of occurrence records in each cell $i$, having area $a_i$, where the cells represent a partition of the study area [@diggleSpatialSpatioTemporalLogGaussian2013;@simpsonGoingGridComputationally2011]. That is they tesselate (but do not overlap) and together cover the entire study area.
Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed to be homogenous across cell $i$) that *would* be seen if there were no spatial bias in rates of reporting. The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\tau_i$ (also assumed homogenous across each cell). Values of $\tau_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates. Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\tau_i$ is not identified in this model, so any model for $\tau_i$ should probably not have an intercept term.

@@

```{r integrated_obs}
p <- 1 - exp(-Lambda[pa_locations])
distribution(y_pa) <- bernoulli(p)
distribution(y_count) <- poisson(alpha_count * Lambda[count_locations])
```
Where `pa_locations` and `count_locations` are numeric indices to the sites with to presence-absence and count data respectively, and `y_pa` and `y_count` are the corresponding vectors of observed data.

@@

Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $M$ life stages (indexed with $j$) at site $i$, at time $t$. $\mathbf{n}_{i,t}$ is evolved through time by multiplication against a timepoint-specific transition matrix $\mathbf{A}_{i,t}$, contructed of timepoint-specific vital rates that depend on the total population density (i.e. across all life stages $M$) in the previous time point $\mathbf{N}_{i,t}$. For example, these density-dependent relationships could be a linear function of population density (controlled by parameters $\alpha_F$ and $\alpha_S$) on the link-scale of the vital rates. Since the transition matrix is no longer temporally-static, the population dynamics cannot be summarised by a population growth rate. Instead, iteration of the matrix using the density-dependent vital rate functions causes the vector $\mathbf{n}_{i,t}$ of population densities to converge on a static value after a sufficient number of iterations, $T$. This terminal density can be used directly as the relative abundance. As in the basic DSDM, the parameters $\alpha_F$ and $\alpha_S$ control both density-dependence and the detection rate, since these two processes cannot be differentiated without population growth data.

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
- `alpha_F`: a scalar for the effect of population density on log-fecundity
- `alpha_S`: a scalar for the effect of population density on logit-survival
- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles

@@

Lambda <- iterations$stable_population
```
Given additional scalar parameters `alpha_F` and `alpha_S`.

@@

Given temporally-indexed response data $y_{i,t}$, temporal variation in environmental conditions and therefore vital rates can be included in a DSDM in one of two ways. The applicability of each of these depends on both the temporal scale of the data (the period of time betwen the observations) and the demographics of the species.

If the length of time between observations is on the order of multiple generations opf the species population, it may be reasonable to assume that the observed population sixes are reflective of contemporary environmental conditions only, and not past fluctuations in environental conditions. In other words, that the population is at, or close to, the long-term carrying capacity of the species if the current environmental conditions were to be held static. In this situation, considering temporal change in environmental conditions is simply a matter of redifining the index $i$ in the basic DSDM to indeex unique combinations of time and site, instead of just site. The relative abundance parameter for each of these observations is therefore computed from the intrinsic growth rate of the transition matrix at the corresponding environmental conditions.

@@

Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive timesteps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above. Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over timesteps $t$ corresponding to time points in the dataset (and all intermediate timesteps). Thus the population at each timestep will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.

@@

- `X_list`: a list of length $T$ of $N$x$K$ design matrices for each of the time points
- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
- `timesteps`: the value of $T$.
- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
- `alpha_F`: a scalar for the effect of population density on log-fecundity
- `alpha_S`: a scalar for the effect of population density on logit-survival
- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
- `y`: an $N$x$T$ matrix of presence-absence data at each timestep and site

@@

@@

Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix that has the population transition matrices for each species on the diagonal. For model of $L$ species each with $M$ lifestages $\mathbf{A}_i$ therefore has dimension $LM$x$LM$. The off-diagonal blocks are all zero-matrices, since an individual can neither change species nor produce offspring of another species. Interactions between species can be encoded via density-dependent vital rates, with the density of both conspecific and intraspecific populations affecting the vital rates of each species.

As in the single-species density-dependent vital rate model above the relative population density for each species is computed by iteration; at each timestep recalculating the transition matrix from population densities of all species and matrix-multiplying it by the current population density vector to obtain the subsequent population density vector. The transition matrix $\mathbf{A}_i$ is sparse; with only 1 in $L$ elements being non-zero. This iteration can therefore be carried out efficiently with sparse matrix-multiplication operations.

@@

where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are 3D arrays with dimension $L$x$N$x$T$ giving the fecundity, adult and juvenile survival and relative population density of each species $l$ at site $i$ at time $t$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ or regression coefficients for the submodels for each of species, and $L$x$L$ matrices $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$ of density-dependence effects between specie. I.e. each element $\mathbf{\alpha}_{F_{l,l'}}$ gives the effect of the density of the population of species $l'$ on the log-fecundity of species $l$. Off-diagonal elements of $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$  quantify fecundity- and survival-mediated interactions between species, whilst the diagonal elements quantify conspecific density-dependence.

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a $K$x$L$ matrix of regression coefficients for log-fecundity for each species
- `beta_S`: a $K$x$L$ matrix of regression coefficients for logit-survival for each species
- `alpha_F`: a $K$x$K$ matrix of the intra- and inter-specific effects of population density on log-fecundity
- `alpha_S`: a $K$x$K$ matrix of the intra- and inter-specific effects of population density on logit-survival
- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival betwen adults and juveniles for each species

@@

Similarly to the biotic interactions model, we can consider a population of single species but with $L$ different genotypes with a blocked transition matrix. In this model, the genotypes are linked via reproduction rather than density-dependent vital rates. tThe $L$x$L$ matrix $\mathbf{m}$ quantifies mutation rates between different genotypes. For an individual of genotype $l$, each of its offspring has probability $\mathbf{m}_{l,l'}$ of being of genotype $l'$. Each genotype has its own vital rate submodels, representing a linked phenotype driving population dynamics for that genotype. By analysing the combined transition matrix $\mathbf{A}^*_i$ at each site $i$ with different environmental covariates, the model can consider how selection for each genotype varies across the environment, and predict the genetic makeup of populations.

@@

where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are $L$x$N$ matrices giving the fecundity, adult and juvenile survival and relative population density of each genotype $l$ at site $i$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ of regression coefficients for the submodels for each of species. $\mathbf{m}$ is an $L$x$L$ matrix of positive values with columns summing to one, giving the probabilities of a parent of one genotype producing offspring of each genotype.

@@

where $\mathbf{B}$ is an $M$x$M$ matrix with ones on the first row and zeros elsewhere (ie. a mask indicating cells representing fecundity rates), $\mathbf{C}$ is its reverse (zeros on the first row, ones elsewhere) $\mathbf{1}$ is length $L$ vector of ones, $\mathbf{I}$ is an identity matrix of dimension $L$, and $\odot$ and $\otimes$ represent the Hadamard (elementwise) and Kronecker products respectively.

@@

```{r evolutionary_dynamics}

```

In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$. To consider interacting population dynamics across sites, we instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure. The the $i$^{th} block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this timestep. $d_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area and ensure $d_{j,i,i'}$ is a valid probability.

By connecting the populations at all sites with immigration and emigration potential, we now have a single intrinsic growth rate for the entire population rather than for th eopulation at each site. Rather than computing the relative abundance at each site $\Lambda_i$ as a function of an intrinsic growth rate, we can instead sum stable stage distributions over the stages in each site to yeild a stable population distribution over sites. $\Lambda_i$ can therefore be computed as proportional (with constant $\alpha$) to this relative population size.

- `A`: a 3D $N$x$M$x$M$ array giving the population transition matrix at each site
- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence

Where the transition matrix is assumed to be dynamic - ie. to model the effects of environmental stochasticity, depsnity-dependent vital rates, or population time series - we use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above. Because we explicitly iterate the matrix, we can apply dispersal with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix. As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.

I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each lifestage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps. For simplicity of notation, we assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.

where $\mathbf{D}$ is an $N$x$N$ matrix of distances between sites, used to define $N$x$N$ matrix $\mathbf{d}_j$ of dispersal probabilitiesfor life stage $j$, which sums to 1 across each column.

- `A_t_fun`: a function yielding for each time point a $N$x$M$x$M$ array of transition matrices for each site
- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
- `timesteps`: the number of timesteps to iterate for
- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
- `...`: additional named arguments to `A_t_fun`, which may be lists with different elements to be used at each time point

>>>>

Each section below details a modelling option for a specific component of the DSDM - the observation model, abundance model, or population dynamic model. These components are largely inter-operable; a user can pick one modelling option from each category that best fits the ecology of the species of interest and the ecological question at hand. The one exception to this is that abundance models using as input the intrinsic growth rate of the species (the Beverton-Holt model in the manuscript example, and the Ricker model introduced here) cannot be used in combination with the population dynamic models that have either dynamic transition matrices or dispersal, since those population dynamic models do not yield a separate intrinsic growth rate for each site. With those models, the relative abundance is instead calculated directly from the dynamics, so these abundance models are not needed.

This appendix only considers the structures of the proposed models, specification of priors for parameters in the models are not discussed here. See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta. I reiterate here that the aim of DSDMs is to incorporate ecological knowledge via informative priors, and the parameters of the proposed models are unlikely to be identified with uninformative priors.

@@

For ease of reference, I repeat here the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript. It presents the model in its three parts, each of which can be modified independently. 
## Observation model

@@

@@

- `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence

@@

@@

The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous time step (columns) to each stage in the present time step (rows). Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult. For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles

@@

@@

Below I provide alternative observation models for count, and presence-only data. occupancy detection observation models can be implemented using other greta extension packages, but are beyond the scope of this manuscript.

@@

The relative abundance variable $Lambda_i$ is the parameter of the Poisson distribution, so a Poisson count model is simply given by:

@@

Where count data are over-dispersed with respect to the expected counts $\Lambda_i$, this over-dispersion can be modelled with a negative binomial distribution as:

@@

- `size`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution
- `Lambda`: a (positive) vector of the relative abundances for each observation
- `y`: a vector of (non-negative) integers encoding count observations

```{r negbin_obs}
prob <- size / (size + Lambda)
distribution(y) <- negative_binomial(size, prob)
```

@@

Presence-only data - where records consist of locations where the species has been reported as being present - can be considered as a thinned, inhomogenous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015]. The likelihood for this model comprises an intractable integral over the study area, which can be approximated in a number of ways that reduce the model to having a Poisson observation model. Here I apply the piece-wise constant approximation, defining the observations $y_i$ to be counts of occurrence records in each cell $i$, having area $a_i$, where the cells represent a partition of the study area [@diggleSpatialSpatioTemporalLogGaussian2013;@simpsonGoingGridComputationally2011]. That is they tessellate (but do not overlap) and together cover the entire study area.
Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed to be homogeneous across cell $i$) that *would* be seen if there were no spatial bias in rates of reporting. The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\tau_i$ (also assumed homogeneous across each cell). Values of $\tau_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression sub-model against detection covariates. Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\tau_i$ is not identified in this model, so any model for $\tau_i$ should probably not have an intercept term.

@@

```{r integrated_obs}
p <- 1 - exp(-Lambda[pa_locations])
distribution(y_pa) <- bernoulli(p)
distribution(y_count) <- poisson(alpha_count * Lambda[count_locations])
```

@@

Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $M$ life stages (indexed with $j$) at site $i$, at time $t$. $\mathbf{n}_{i,t}$ is evolved through time by multiplication against a time-point-specific transition matrix $\mathbf{A}_{i,t}$, constructed of time-point-specific vital rates that depend on the total population density (i.e. across all life stages $M$) in the previous time point $\mathbf{N}_{i,t}$. For example, these density-dependent relationships could be a linear function of population density (controlled by parameters $\alpha_F$ and $\alpha_S$) on the link-scale of the vital rates. Since the transition matrix is no longer temporally-static, the population dynamics cannot be summarised by a population growth rate. Instead, iteration of the matrix using the density-dependent vital rate functions causes the vector $\mathbf{n}_{i,t}$ of population densities to converge on a static value after a sufficient number of iterations, $T$. This terminal density can be used directly as the relative abundance. As in the basic DSDM, the parameters $\alpha_F$ and $\alpha_S$ control both density-dependence and the detection rate, since these two processes cannot be differentiated without population growth data.

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
- `alpha_F`: a scalar for the effect of population density on log-fecundity
- `alpha_S`: a scalar for the effect of population density on logit-survival
- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles

@@

@@

Given temporally-indexed response data $y_{i,t}$, temporal variation in environmental conditions and therefore vital rates can be included in a DSDM in one of two ways. The applicability of each of these depends on both the temporal scale of the data (the period of time between the observations) and the demographics of the species.

If the length of time between observations is on the order of multiple generations of the species population, it may be reasonable to assume that the observed population sixes are reflective of contemporary environmental conditions only, and not past fluctuations in environmental conditions. In other words, that the population is at, or close to, the long-term carrying capacity of the species if the current environmental conditions were to be held static. In this situation, considering temporal change in environmental conditions is simply a matter of redefining the index $i$ in the basic DSDM to index unique combinations of time and site, instead of just site. The relative abundance parameter for each of these observations is therefore computed from the intrinsic growth rate of the transition matrix at the corresponding environmental conditions.

@@

Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive time-steps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above. Instead of iterating for some arbitrary number of iterations $T$ and taking the final state, instead the population is iterated over time-steps $t$ corresponding to time points in the dataset (and all intermediate time-steps). Thus the population at each time-step will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.

@@

- `X_list`: a list of length $T$ of $N$x$K$ design matrices for each of the time points
- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
- `timesteps`: the value of $T$.
- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
- `alpha_F`: a scalar for the effect of population density on log-fecundity
- `alpha_S`: a scalar for the effect of population density on logit-survival
- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
- `y`: an $N$x$T$ matrix of presence-absence data at each time-step and site

@@

In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$. To consider interacting population dynamics across sites, we instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure. The the $i^{th}$ block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this time-step. $d_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area and ensure $d_{j,i,i'}$ is a valid probability.

By connecting the populations at all sites with immigration and emigration potential, we now have a single intrinsic growth rate for the entire population rather than for the population at each site. Rather than computing the relative abundance at each site $\Lambda_i$ as a function of an intrinsic growth rate, we can instead sum stable stage distributions over the stages in each site to yield a stable population distribution over sites. $\Lambda_i$ can therefore be computed as proportional (with constant $\alpha$) to this relative population size.

- `A`: a 3D $N$x$M$x$M$ array giving the population transition matrix at each site
- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence

Where the transition matrix is assumed to be dynamic - i.e.. to model the effects of environmental stochasticity, density-dependent vital rates, or population time series - we use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above. Because we explicitly iterate the matrix, we can apply dispersal with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix. As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.

I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each life-stage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps. For simplicity of notation, we assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.

where $\mathbf{D}$ is an $N$x$N$ matrix of distances between sites, used to define $N$x$N$ matrix $\mathbf{d}_j$ of dispersal probabilities for life stage $j$, which sums to 1 across each column.

- `A_t_fun`: a function yielding for each time point a $N$x$M$x$M$ array of transition matrices for each site
- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
- `timesteps`: the number of time-steps to iterate for
- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
- `...`: additional named arguments to `A_t_fun`, which may be lists with different elements to be used at each time point

@@

Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix that has the population transition matrices for each species on the diagonal. For model of $L$ species each with $M$ life-stages $\mathbf{A}_i$ therefore has dimension $LM$x$LM$. The off-diagonal blocks are all zero-matrices, since an individual can neither change species nor produce offspring of another species. Interactions between species can be encoded via density-dependent vital rates, with the density of both conspecific and intraspecific populations affecting the vital rates of each species.

As in the single-species density-dependent vital rate model above the relative population density for each species is computed by iteration; at each time-step recalculating the transition matrix from population densities of all species and matrix-multiplying it by the current population density vector to obtain the subsequent population density vector. The transition matrix $\mathbf{A}_i$ is sparse; with only 1 in $L$ elements being non-zero. This iteration can therefore be carried out efficiently with sparse matrix-multiplication operations.

@@

where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are 3D arrays with dimension $L$x$N$x$T$ giving the fecundity, adult and juvenile survival and relative population density of each species $l$ at site $i$ at time $t$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ or regression coefficients for the sub-models for each of species, and $L$x$L$ matrices $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$ of density-dependence effects between specie. I.e. each element $\mathbf{\alpha}_{F_{l,l'}}$ gives the effect of the density of the population of species $l'$ on the log-fecundity of species $l$. Off-diagonal elements of $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$  quantify fecundity- and survival-mediated interactions between species, whilst the diagonal elements quantify conspecific density-dependence.

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a $K$x$L$ matrix of regression coefficients for log-fecundity for each species
- `beta_S`: a $K$x$L$ matrix of regression coefficients for logit-survival for each species
- `alpha_F`: a $L$x$L$ matrix of the intra- and inter-specific effects of population density on log-fecundity
- `alpha_S`: a $L$x$L$ matrix of the intra- and inter-specific effects of population density on logit-survival
- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival between adults and juveniles for each species

@@

Similarly to the biotic interactions model, we can consider a population of single species but with $L$ different genotypes with a blocked transition matrix. In this model, the genotypes are linked via reproduction rather than density-dependent vital rates. The $L$x$L$ matrix $\mathbf{m}$ quantifies mutation rates between different genotypes. For an individual of genotype $l$, each of its offspring has probability $\mathbf{m}_{l,l'}$ of being of genotype $l'$. Each genotype has its own vital rate sub-models, representing a linked phenotype driving population dynamics for that genotype. By analysing the combined transition matrix $\mathbf{A}^*_i$ at each site $i$ with different environmental covariates, the model can consider how selection for each genotype varies across the environment, and predict the genetic makeup of populations.

@@

where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are $L$x$N$ matrices giving the fecundity, adult and juvenile survival and relative population density of each genotype $l$ at site $i$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ of regression coefficients for the sub-models for each of species. $\mathbf{m}$ is an $L$x$L$ matrix of positive values with columns summing to one, giving the probabilities of a parent of one genotype producing offspring of each genotype.

@@

where $\mathbf{B}$ is an $M$x$M$ matrix with ones on the first row and zeros elsewhere (i.e.. a mask indicating cells representing fecundity rates), $\mathbf{C}$ is its reverse (zeros on the first row, ones elsewhere) $\mathbf{1}$ is length $L$ vector of ones, $\mathbf{I}$ is an identity matrix of dimension $L$, and $\odot$ and $\otimes$ represent the Hadamard (element-wise) and Kronecker products respectively.

@@

- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
- `beta_F`: a $K$x$L$ matrix of regression coefficients for log-fecundity for each genotype
- `beta_S`: a $K$x$L$ matrix of regression coefficients for logit-survival for each genotype
- `m`: a $L$x$L$ matrix of the transition (mutation) rates between genotypes, with columns summing to one
- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival between adults and juveniles for each species

```{r evolutionary_dynamics}
# vital rate models
log_F<- X %*% beta_F
logit_S<- X %*% beta_S
F <- exp(log_F)
S_A <- plogis(logit_S)
S_J <- plogis(logit_S - gamma)

L <- length(gamma)
N <- nrow(X)
M <- 2
padding <- zeros(N, L * M)
B <- zeros(M, M)
B[1, ] <- 1
C <- 1 - B
dd_weights <- diag(L)[rep(1:L, each = M), ]

# array A_star of transition matrices
mutation <- kronecker(m, B) + kronecker(diag(L), C)

F_rep <- do.call(rbind, replicate(M, F, simplify = FALSE))
dim(F_rep) <- c(N, L * M)
F_kron_tB <- abind(F_rep, padding, along = 3)

SJ_rep <- do.call(cbind, replicate(M, S_J, simplify = FALSE))
SJ_kron_B <- rbind(SJ_rep, padding)
dim(SJ_kron_B) <- c(N, L * 2,  M)

SA_rep <- do.call(cbind, replicate(M, S_A, simplify = FALSE))
SA_kron_C <- rbind(padding, SA_rep)
dim(SA_kron_C) <- c(N, L * 2,  M)

R <- F_kron_tB + SJ_kron_B + SA_kron_C
one_kron_R <- do.call(abind, c(replicate(L, R, simplify = FALSE), list(along = 3)))
dim(one_kron_R) <- c(N, L * M, L * M)
one_kron_R <- aperm(one_kron_R, c(1, 3, 2))

# do multiplication across N as a matrix
dim(one_kron_R) <- c(N, L ^ 2 * M ^ 2)
dim(mutation) <- c(L ^ 2 * M ^ 2)
A_star <- sweep(one_kron_R, 2, mutation, "*")
dim(A_star) <- c(N, L * M, L * M)

analysis <- iterate_matrix(A_star)
v_star <- analysis$stable_abundance_distribution
v_standardised <- analysis$stable_abundance_distribution %*% dd_weights
lambda <- analysis$lambda
K <- (lambda - 1) * alpha
Lambda_star <- log1pe(K)
Lambda <- Lambda_star * v_standardised
```
